---
title: "Lab 5 : Data wrangling and graphing COVID-19 reporting data (Part 2)"
output:
  html_document:
    toc: true
    toc_depth : 4
    theme: united
    highlight: tango
editor_options: 
  chunk_output_type: console
---

*I recognize, and fully understand, that this data maybe emotionally difficult to work. My intention is to make these lab relevant, allowing you to gather your own insights directly from new visualizations of the data. Please let me know if you would rather not work with the data.*

## Learning Objectives

* Understanding the sources of SARS-CoV-2 incidence reports
* Accessing data remotely
* Joining tables together
* Wide and long table formats
* More data visualization with ggpl02

### Joining tables

[Tidyverse Cookbook](https://rstudio-education.github.io/tidyverse-cookbook/) has many practical solutions to problems. We will go over different ways to [join tables](https://rstudio-education.github.io/tidyverse-cookbook/transform-tables.html#joins). 

Let's start by loading `tidyverse`

```{r}
library(tidyverse)
```

```{r, message=FALSE}
Confirmed_State_3_13 <-   read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_daily_reports/03-13-2020.csv")) %>%
  rename(Country_Region = "Country/Region", Province_State = "Province/State") %>% 
  filter (Country_Region == "US") %>% 
  group_by(Province_State, Country_Region) %>% 
  summarise(Confirmed = sum(Confirmed)) 
```

```{r}
str(Confirmed_State_3_13)
```

```{r, message=FALSE}
Confirmed_State_9_13 <-   read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_daily_reports/09-13-2020.csv")) %>% 
  filter (Country_Region == "US") %>% 
  group_by(Province_State, Country_Region) %>% 
  summarise(Confirmed = sum(Confirmed)) 
```

```{r}
str(Confirmed_State_9_13)
```

Notice from the above `str' calls that there are 53 states in the table on 3/13/2020 and 58 states in the table on 9/13/2020

```{r}
setdiff(Confirmed_State_9_13$Province_State, Confirmed_State_3_13$Province_State)
```

There are 4 US territories a field you may have notice last week "Recovered". We don't need the "Recovered" row so let's filter it from the data set

```{r}
Confirmed_State_9_13 <- Confirmed_State_9_13 %>% 
  filter(Province_State != "Recovered") 
```

Since all of the states and territories in 9/13/2020 are found in 3/13/2020 a `left_join` or `full_join` can be used to merge the tables

```{r}
Confirmed_State_3_13_9_13_joined <- full_join(Confirmed_State_3_13,
      Confirmed_State_9_13, by = c("Province_State"))
head(Confirmed_State_3_13_9_13_joined)
```

This results in the values for Guam, Northern Mariana Islands, Puerto Rico and the Virgin Islands being NA for 3/13

```{r}
tail(Confirmed_State_3_13_9_13_joined, 5)
```

or

```{r}
which(is.na(Confirmed_State_3_13_9_13_joined))
```

Let's turn the`NA`s into 0. Also notice that we know have columns `Confirmed.x` `Country_Region.x`, `Confirmed.y` `Country_Region.y`. So let's rename the `Confirmed` columns and delete the `Country_Region` columns

```{r}
Confirmed_State_3_13_9_13_joined <- full_join(Confirmed_State_3_13,
      Confirmed_State_9_13, by = c("Province_State")) %>% 
      rename(Confirmed_3_13_2020 = "Confirmed.x", Confirmed_9_13_2020 = "Confirmed.y") %>% 
      select(-Country_Region.x, -Country_Region.y) %>% 
      replace_na(list(Confirmed_3_13_2020 = 0))
head(Confirmed_State_3_13_9_13_joined)
```

```{r}
which(is.na(Confirmed_State_3_13_9_13_joined))
```

### Switching between wide and long table formats.

`gather` and `spead` were used in the Data Carpentries tutorial to switch between wide and long table formats. They work fine, but are being replaced by the newer functions `pivot_longer` and `pivot_wider`. which we will use below. See the [pivot vignette]( https://tidyr.tidyverse.org/articles/pivot.html) for other examples.


```{r}
Confirmed_State_3_13_9_13_joined_long <- Confirmed_State_3_13_9_13_joined %>% 
              pivot_longer(-c(Province_State),
                            names_to = "Date", values_to = "Confirmed")
```


Now we can use this table in plot that shows the confirmed cases for both dates.

```{r, fig.height = 10}
# to adjust figure size {r, fig.width = 5, fig.height = 10}
ggplot(Confirmed_State_3_13_9_13_joined_long, aes(x = Confirmed,  y = Province_State))  + 
    geom_point(aes(color = Date)) 
```

### Working with the time series data

#### Data Wrangling

Start by loading the most recent times series data for confirmed cases

```{r, message=FALSE}
time_series_confirmed <- read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv")) %>%
  rename(Province_State = "Province/State", Country_Region = "Country/Region")
```

Note that in some cases it is best to have the date on your computer and then load the file into R.  

```{r, eval=FALSE}
 download.file(url="https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv", 
               destfile = "data/time_series_covid19_confirmed_global.csv")
```

```{r}
time_series_confirmed <- read_csv("data/time_series_covid19_confirmed_global.csv")%>%
  rename(Province_State = "Province/State", Country_Region = "Country/Region")
```

Check the table properties to make sure the data imported as we expected

```{r}
head(time_series_confirmed)
```

As noted above this data is in wide format. To convert to long format

```{r}
time_series_confirmed_long <- time_series_confirmed %>% 
               pivot_longer(-c(Province_State, Country_Region, Lat, Long),
                            names_to = "Date", values_to = "Confirmed") 
```

Let's look at the format of the data frame (tibble) now

```{r}
head(time_series_confirmed_long)
```

It would be convenient to have the confirmed cases and deaths in the same table. We can create another table with the deaths and then join the two tables. First let's 

```{r, eval=FALSE}
 download.file(url="https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv", 
               destfile = "data/time_series_covid19_deaths_global.csv")
```

```{r}
time_series_deaths <- read_csv("data/time_series_covid19_deaths_global.csv")%>%
  rename(Province_State = "Province/State", Country_Region = "Country/Region")
```

As noted above this data is in wide format. To convert to long format

```{r}
time_series_deaths_long <- time_series_deaths %>% 
               pivot_longer(-c(Province_State, Country_Region, Lat, Long),
                            names_to = "Date", values_to = "Deaths") 
head(time_series_deaths_long)
```


#### Joining the time series tables

To join 2 tables we need a common column in which each row is a unique name. This is often called the Key. None of the columns in this data set meet that criteria, but we can create a column to serve as our key by mergering the names for the Province_State, Country_Region and Date columns using __unite__

```{r}
time_series_confirmed_long <- time_series_confirmed_long %>% 
  unite(Key, Province_State, Country_Region, Date, sep = ".", remove = FALSE)
head(time_series_confirmed_long)
```


For the second table we can do the same and eliminate the columns that are redundant so that we just have the Key and Deaths columns

```{r}
time_series_deaths_long <- time_series_deaths_long %>% 
  unite(Key, Province_State, Country_Region, Date, sep = ".") %>% 
  select(Key, Deaths)
```

Now to join the tables. A __full_join__ which means that the keys that are common to the tables will be joined and there will be counts for both Confirmed and Deaths. Where there are not matching values, returns NA for the one missing. __select__ can be used to remove the key after the join since we don't have further need for it.

```{r}
time_series_long_joined <- full_join(time_series_confirmed_long,
              time_series_deaths_long, by = c("Key")) %>% 
              select(-Key)
head(time_series_long_joined)
```

Check to make sure neither the Cofirmed or Death counts have NA as a value (both tables have the same number of rows, but we should check to make sure)

```{r}
which(is.na(time_series_long_joined$Confirmed))
which(is.na(time_series_long_joined$Deaths))
```

There are no NA values. Now we are in good shape to proceed to using these table for graphing, but first reformat Date

```{r}
library(lubridate)
time_series_long_joined$Date <- mdy(time_series_long_joined$Date)
```

Confirmed and Deaths are count data and there may be times when we want to plot both on the same graph. There are multiple ways to accomplish this and one is through creating a column using _pivot_long

```{r}
time_series_long_joined_counts <- time_series_long_joined %>% 
  pivot_longer(-c(Province_State, Country_Region, Lat, Long, Date),
               names_to = "Report_Type", values_to = "Counts")
head(time_series_long_joined_counts)
```

#### Making Graphs from the time series data

To make a times series graph of the confirmed cases we need to summarize the Country date to count up the individual state data for the US. 

```{r}
time_series_long_joined %>% 
  group_by(Country_Region,Date) %>% 
  summarise_at(c("Confirmed", "Deaths"), sum) %>% 
  filter (Country_Region == "US") %>% 
    ggplot(aes(x = Date,  y = Deaths)) + 
    geom_point() +
    geom_line() +
    ggtitle("US COVID-19 Deaths")
```

Let's look at the US data in the context of a few other countries

```{r}
time_series_long_joined %>% 
  group_by(Country_Region,Date) %>% 
  summarise_at(c("Confirmed", "Deaths"), sum) %>% 
  filter (Country_Region %in% c("China","Japan", "Korea, South",
                                "Italy","Spain", "US")) %>% 
    ggplot(aes(x = Date,  y = Deaths)) + 
    geom_point() +
    geom_line() +
    ggtitle("COVID-19 Deaths") +
    facet_wrap(~Country_Region, ncol=2, scales="free_y")
```

Now several countries on the same graph

```{r}
time_series_long_joined %>% 
    group_by(Country_Region,Date) %>% 
    summarise_at(c("Confirmed", "Deaths"), sum) %>% 
    filter (Country_Region %in% c("China","France","Italy", 
                                "Korea, South", "US")) %>% 
    ggplot(aes(x = Date,  y = Deaths, color = Country_Region)) + 
    geom_point() +
    geom_line() +
    ggtitle("COVID-19 Deaths")
```

We can use the alternative data frame with column Report_Type to show Confirmed and Deaths. Because the counts for Deaths are low relative to Confirmed a log scale works best here for seeing the pattern

```{r}
time_series_long_joined_counts %>% 
  group_by(Country_Region, Report_Type, Date) %>% 
  summarise(Counts = sum(Counts)) %>% 
  filter (Country_Region == "US") %>% 
    ggplot(aes(x = Date,  y = log2(Counts), fill = Report_Type, color = Report_Type)) + 
    geom_point() +
    geom_line() +
    ggtitle("US COVID-19 Cases")
```

## Exercises

1. Finish the [Data Visualization tutorial](https://datacarpentry.org/R-ecology-lesson/04-visualization-ggplot2.html#Boxplot) and add it your GitPage web site

2. Revise the above example for joining tables using 6/13/2020 and 9/13/2020. Plot the data as a bar plot. Make sure your graph looks nice in the report by adjusting the height of the graph in the R code chunk header.

3. Add more informative x and y labels and a title to the above graph in Ex2.

4. Using the time series data. Make a plot of the total number of confirmed deaths per day worldwide 

5. Use Mutate to make a new column with deaths/confirmed cases (Hint: this might be best done after joining the tables, but before pivoting longer).

6. Plot US deaths/confirmed cases per day.

7. Make a single graph with from the times series data with the 10 countries with the hightest death total.

8. Using Facet_Wraps make a graph from the times series data with the 10 countries with the hightest death total. Format the table to look nice in the report by changing the number of rows in your graph and/or adjusting the height in the R code chunk header.

9. Using Facet_Wraps make a graph from the times series data for confirmed cases for all US
states and territories. Format the table to look nice in the report by changing the number of rows in your graph and/or adjusting the height in the R code chunk header.

10. Apply a theme other than the one used in the Data Carpentries tutorial to the graph in Ex8. 



